---
title: "Biostatistics II 2020 - Midterm Assignment (Table 1)"
author: "Vinyas Harish, MD/PhD student year 3, University of Toronto"
date: "Feb 29 2020"
output: html_document
---

# Description

This R markdown script is used to create table ones for the midterm assignment. 

Dataset Description: Dataset of 560k patient visits in three Yale ERs.

Dataset Source: https://github.com/yaleemmlc/admissionprediction (Github, can download the dataframe), 
https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0201016#pone.0201016.ref012(accompanying PLoS ONE paper)

## Load libraries and dataframes
```{r}
library(knitr)
library(table1)
library(tidyverse)

load("~/Desktop/dispPred.RDa")
load("~/Desktop/rxPred.RDa")

```

## More cleaning that came up, primary outcome
```{r, eval=FALSE}
#Recat race to fold more groups into other
levels(dispositionPrediction_wCC$race) <- list("White or Caucasian" = c("White or Caucasian"),
                                              "Black or African American " = c("Black or African American"),
                                              "Hispanic or Latino" = c ("Hispanic or Latino"),
                                              "Other" = c(" American Indian or Alaska Native","Asian","Native Hawaiian or Other Pacific                                                                   Islander"),
                                              "Unknown" = c("Patient Refused", "Unknown"))
dispositionPrediction_wCC$race[is.na(dispositionPrediction_wCC$race)] <- "Unknown" #Recat missing race

#Rename hospital deps
levels(dispositionPrediction_wCC$dep_name) <- list("Level 1 Trauma Center" = c("A"),
                                                   "Community Hospital" = c("B"),
                                                   "Freestanding ER" = c("C"))

levels(dispositionPrediction_wCC$arrivalmode) <- list("Emergency Services" = c("ambulance", "police"),
                                                      "Car" = c("Car"),
                                                      "Walk-in" = c ("Walk-in", "Public Transportation", "Wheelchair"),
                                                      "Other" = c("Other"))

levels(dispositionPrediction_wCC$previousdispo) <- list("Admit" = c("Admit", "AMA", "Eloped", "Observation", "Send to L&D", "Transfer to Another Facility"),
                                                        "Discharge" = c("Discharge"),
                                                        "No previous disposition" = c("No previous dispo", "LWBS before Triage","LWBS after Triage"))

dispositionPrediction_wCC <- dispositionPrediction_wCC[!is.na(dispositionPrediction_wCC$arrivalmode),]

dispositionPrediction_wCC <- dispositionPrediction_wCC[!is.na(dispositionPrediction_wCC$cc_const),] #assuming this takes care of everyone

collapse_cc <- function(dat){
  dat$ptx <- seq.int(nrow(dat))
  
  dat <- dat %>% 
    gather(contains('cc'), key='chief_complaint', value='n_complaints') %>% 
    group_by(ptx) %>% 
    mutate(rank = rank(n_complaints, ties.method = 'first')) %>% 
    top_n(n=1, wt=rank) %>% 
    select(ptx, chief_complaint, n_complaints, everything()) %>% 
    ungroup(ptx) %>% 
    arrange(ptx) %>% 
    select(-ptx, -rank)
  
  return(dat)
}


dispPred <- dispositionPrediction_wCC %>% collapse_cc(.)
```

## More cleaning that came up, secondary outcome
```{r, eval=FALSE}
#Recat race to fold more groups into other
levels(rxPred_wCC$race) <- list("White or Caucasian" = c("White or Caucasian"),
                                              "Black or African American " = c("Black or African American"),
                                              "Hispanic or Latino" = c ("Hispanic or Latino"),
                                              "Other" = c(" American Indian or Alaska Native","Asian","Native Hawaiian or Other Pacific                                                                   Islander"),
                                              "Unknown" = c("Patient Refused", "Unknown"))
rxPred_wCC$race[is.na(rxPred_wCC$race)] <- "Unknown" #Recat missing race

#Rename hospital deps
levels(rxPred_wCC$dep_name) <- list("Level 1 Trauma Center" = c("A"),
                                                   "Community Hospital" = c("B"),
                                                   "Freestanding ER" = c("C"))

levels(rxPred_wCC$arrivalmode) <- list("Emergency Services" = c("ambulance", "police"),
                                                      "Car" = c("Car"),
                                                      "Walk-in" = c ("Walk-in", "Public Transportation", "Wheelchair"),
                                                      "Other" = c("Other"))

levels(rxPred_wCC$previousdispo) <- list("Admit" = c("Admit", "AMA", "Eloped", "Observation", "Send to L&D", "Transfer to Another Facility"),
                                                        "Discharge" = c("Discharge"),
                                                        "No previous disposition" = c("No previous dispo", "LWBS before Triage","LWBS after Triage"))

rxPred_wCC <- rxPred_wCC[!is.na(rxPred_wCC$arrivalmode),]

rxPred_wCC <- rxPred_wCC[!is.na(rxPred_wCC$cc_const),] #assuming this takes care of everyone

rxPred <- rxPred_wCC %>% collapse_cc(.)
```

## Cleanup Workspace 
```{r,eval=FALSE}
#Update saved dataframes
save(dispPred, file="dispPred.Rda")
save(rxPred, file="rxPred.Rda")
write.csv(dispPred,'dispPred.csv')
write.csv(rxPred, 'rxPred.csv')

#Clean up old files
rm(rxPred_wCC)
rm(dispositionPrediction_wCC)

#Get datatypes for each column in dataframe (they should be identical)
datatypes <- sapply(dispPred, class)
print(datatypes) #All factors or numeric which is convenient for this assignment

```


##Create Table 1a. Primary Outcome (Admission)
```{r}
#Group by: pt demographics, triage vitals (including ESI), chief complaint, and past hospital usage
table1(~ factor(gender) + age + factor(race) + factor (lang) + factor (religion) + factor(maritalstatus) + factor(employstatus)
       + factor(insurance_status) + dep_name + factor(arrivalmode) + factor(arrivalday) + factor(arrivaltime) + triage_vital_hr +
      + triage_vital_sbp + triage_vital_dbp + triage_vital_rr + triage_vital_o2 + triage_vital_temp + esi + chief_complaint + n_complaints + previousdispo + n_edvisits +n_admissions + n_surgeries| disposition, data=dispPred, render.continuous = c(.="Mean (SD)",.="Median [Q1, Q3]"))

#Statistical tests
testGender = chisq.test(table(dispPred$gender, dispPred$gender))
testAge = wilcox.test(x = table(dispPred$age, dispPred$disposition)) #This is different than SAS

testAge = wilcox.test(x =dispPred$age[which(dispPred$disposition=='Admit' )], 
                      y =dispPred$age[which(dispPred$disposition=='Discharge' )],
                      paired = FALSE, correct = TRUE) #Still different than SAS but closer test stat

testLang = chisq.test(table(dispPred$lang, dispPred$disposition))


testHR = wilcox.test(x =dispPred$triage_vital_hr[which(dispPred$disposition=='Admit' )], 
                      y =dispPred$triage_vital_hr[which(dispPred$disposition=='Discharge' )],
                      paired = FALSE, correct = TRUE) 

testSBP = wilcox.test(x =dispPred$triage_vital_sbp[which(dispPred$disposition=='Admit' )], 
                      y =dispPred$triage_vital_sbp[which(dispPred$disposition=='Discharge' )],
                      paired = FALSE, correct = TRUE) 

testDBP = wilcox.test(x =dispPred$triage_vital_dbp[which(dispPred$disposition=='Admit' )], 
                      y =dispPred$triage_vital_dbp[which(dispPred$disposition=='Discharge' )],
                      paired = FALSE, correct = TRUE) 

testRR = wilcox.test(x =dispPred$triage_vital_rr[which(dispPred$disposition=='Admit' )], 
                      y =dispPred$triage_vital_rr[which(dispPred$disposition=='Discharge' )],
                      paired = FALSE, correct = TRUE) 

testO2 = wilcox.test(x =dispPred$triage_vital_o2[which(dispPred$disposition=='Admit' )], 
                      y =dispPred$triage_vital_o2[which(dispPred$disposition=='Discharge' )],
                      paired = FALSE, correct = TRUE) 

testTemp = wilcox.test(x =dispPred$triage_vital_temp[which(dispPred$disposition=='Admit' )], 
                      y =dispPred$triage_vital_temp[which(dispPred$disposition=='Discharge' )],
                      paired = FALSE, correct = TRUE) 
```

##Create Table 1a. Primary Outcome (Polypharmacy)
```{r}
#Group by: pt demographics, triage vitals (including ESI), chief complaint, and past hospital usage
table1(~ factor(gender) + age + factor(race) + factor (lang) + factor (religion) + factor(maritalstatus) + factor(employstatus)
       + factor(insurance_status) + dep_name + factor(arrivalmode) + factor(arrivalday) + factor(arrivaltime) + chief_complaint + n_complaints +     triage_vital_hr + triage_vital_sbp + triage_vital_dbp + triage_vital_rr + triage_vital_o2 + triage_vital_temp + esi + previousdispo 
      + n_edvisits +n_admissions + n_surgeries + n_meds, data=rxPred, render.continuous = c(.="Mean (SD)",.="Median [Q1, Q3]"))

```



